---
title: "Seychelles Cancer Awareness Survey Planning"
author: "L'Escale, Mahe, Seychelles"
date: 27-29 May 2024
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

if(!require(remotes)) install.packages("remotes")
if(!require(fontawesome)) remotes::install_github("rstudio/fontawesome")

tooltip_css <- "font-family:Arial,Helvetica,sans-serif; font-size:12px; font-weight:bold; color:#FFFFFF; padding:5px"

## Sample size calculator function ----
calculate_sample_size <- function(z, p, d) {
  ss <- (z ^ 2 * (p * (1 - p))) / d ^ 2
  
  ceiling(ss)
}

## Finite population correction ----
correct_sample_finite <- function(n, pop) {
  ss <- n / (1 + n / pop)
  ceiling(ss)
}
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

style_mono_light(
  base_color = "#002147",
  #title_slide_background_image = "https://i.guim.co.uk/img/static/sys-images/Guardian/Pix/pictures/2014/7/29/1406640126780/f089869e-2c47-481b-be03-db976b2ec9e1-1024x768.jpeg?width=620&quality=85&auto=format&fit=max&s=fd5278ce1249bf8f11f78a7bc869933e",
  #title_slide_background_size = "cover",
  link_color = "#214700",
  header_font_google = google_font("Fira Sans"),
  text_font_google   = google_font("Fira Sans Condensed"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.2em",
  header_h1_font_size = "50px",
  header_h2_font_size = "40px",
  header_h3_font_size = "30px",
  text_slide_number_font_size = "0.5em",
  footnote_font_size = "0.5em"
)
```

class: inverse, center, middle

## Technical Notes and Discussions

---

# Outline

* General sample size considerations for surveys

* Specific sample size considerations for specific types of surveys

---

# General sample size considerations for surveys

* Sample size will depend on which indicator/s need to be measured

* Sample size will depend on survey design

* Sample size will depend on amount of resources available

---

<!-- background-color: #FFFFFF -->

# Sample size equation for prevalence surveys

* Basic formula:

$$ n ~ = ~ \frac{z ^ 2 ~ \times ~ p(1 - p)}{d ^ 2}  $$

where:

\\(n ~ = ~ \text{sample size}\\)

\\(z ~ = ~ \text{z-score (standard deviation) for confidence interval required}\\)

\\(p ~ = ~ \text{known proportion of the indicator being measured}\\)

\\(d ~ = ~ \text{precision required}\\)

---

# Sample size considerations for prevalence surveys

* To calculate this adequate sample size there is a simple formula

* However it needs some practical issues in selecting values for the assumptions required in the formula

* In some situations, the decision to select the appropriate values for these assumptions are not simple

---

# Sample size considerations for prevalence surveys - z statisic

* $z$ statistic is usually a choice between these values:

```{r, echo = FALSE}
z <- c(1.65, 1.96, 2.58)

pvalue <- c(0.1, 0.05, 0.01)

cint <- c("90%", "95%", "99%")

data.frame(z, pvalue, cint) |>
  knitr::kable(col.names = c("z statistic", "p-value", "confidence intervals"))
```


* For most prevalence surveys (and most studies), a **95% confidence interval** is what we want to aim for

---

# Sample size considerations for prevalence surveys - true prevalence

* Need to use a known value of true prevalence (\\(p\\)) of the indicator being measured

* This can usually be found from a literature review to find similar studies/surveys that measured similar indicators

* If proposed/planned study is so unique and measures indicators not measured before, then a best guess of true prevalence can be used; or,

* Use a true prevalence value that gives the highest possible sample size requirement - this value is **0.50 (50%)**

---

# Sample size considerations for prevalence surveys - precision

* Precision, in general terms, is the amount of "swing" below and above the estimated prevalence within which the true prevalence lies

* Selecting a value for precision (\\(d\\)) to aim for should take into account the assumed or known true prevalence (\\(p\\)).

* Some authors recommended to select a precision of 5% if the prevalence of the disease is going to be between 10% and 90% 

* However, when the assumed prevalence is too small (going to be below 10%) or too high (going to be greater than 90%), the precision of 5% seems to be inappropriate

---

# Effect of varying true prevalence

```{r, echo = FALSE}
z <- 1.96
p <- c(0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.9, 0.95)
d <- 0.05

ss <- calculate_sample_size(z = z, p = p, d = d)

data.frame(z, p, d, ss) |>
  knitr::kable(col.names = c("z-statistic", "true prevalence", "precision", "sample size"))
```

---

# Effect of varying precision

```{r, echo = FALSE}
z <- 1.96
p <- rep(c(0.05, 0.2,  0.5, 0.7, 0.95), 3)
d <- c(rep(0.03, 5), rep(0.05, 5), rep(0.08, 5))

ss <- calculate_sample_size(z = z, p = p, d = d)

data.frame(z, p, d, ss) |>
  dplyr::arrange(p, d) |>
  dplyr::slice(1:9) |>
  knitr::kable(col.names = c("z-statistic", "true prevalence", "precision", "sample size"))
```

---

# Effect of varying precision - continued

```{r, echo = FALSE}
data.frame(z, p, d, ss) |>
  dplyr::arrange(p, d) |>
  dplyr::slice(10:15) |>
  knitr::kable(col.names = c("z-statistic", "true prevalence", "precision", "sample size"))
```

---

# Sample size considerations for prevalence surveys - finite population

* The size of the universe/total population from which sampling is to be done impacts sample size

* A finite population correction can be applied to sample size calculations to make it more appropriate/applicable to a known population size

---

# Finite population correction for sample size

$$ n_{adjusted} ~ = ~ \frac{n}{1 + \frac{n}{pop}} $$
where:

\\(n_{adjusted} ~ = ~ \text{Adjusted sample size}\\)

\\(n ~ = ~ \text{calculated sample size}\\)

\\(pop ~ = ~ \text{population}\\)

---

# Adjusted sample sizes accounting for finite population

```{r, echo = FALSE}
z <- 1.96
p <- c(0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.9, 0.95)
d <- 0.05

ss <- calculate_sample_size(z = z, p = p, d = d)

adj <- correct_sample_finite(n = ss, pop = 100000)

data.frame(z, p, d, ss, adj) |>
  knitr::kable(col.names = c("z-statistic", "true prevalence", "precision", "sample size", "adjusted"))

```

---

class: inverse, center, middle

# Questions?

---

class: inverse, center, middle

# Thank you!
















